---
title: "Accidents in Israel"
author: "Ariel Karlinsky"
date: "11/08/2020"
output: 
  html_document: 
    theme: cerulean
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align="center")
```
# Preface

This document downloads data on:

- Fatalities in traffic accidents in Israel from 1949 to 2019
- Total KM traveled in Israel in selected years (according to data availability)
- Total number of vehicles in Israel

and merges it with publicly available data on the Israeli population from the World Bank and the Penn World Tables. 

The results are plots which show:

- Fatalities from traffic accidents in Israel
- Rate of fatalities from traffic accidents in Israel (per 100,000 people)
- Rate of fatalities from traffic accidents in Israel (per 1,000,000,000 KM traveled)
- Rate of fatalities from traffic accidents in Israel (per 100,000 KM vehicles)


# Loading and wrangling data

Loading needed packages and setting the plot theme:
```{r echo=T, message=FALSE, warning=FALSE, results='hide'}
pacman::p_load(tidyverse, scales, WDI, pwt9, readxl, plotly, ggpmisc, patchwork)

ggplottheme <- ggpubr::theme_classic2() + 
    theme(
          panel.background = element_rect(fill = "transparent",colour = NA),
          plot.background = element_rect(fill = "transparent",colour = NA),
          plot.title = element_text(hjust = 1),
          plot.subtitle = element_text(hjust = 1),
          plot.caption = element_text(hjust = 1),
          axis.ticks = element_blank(),
          strip.background = element_blank(),
          panel.spacing = unit(2, "lines")
    )
theme_set(ggplottheme)

```

Download the data from the CBS:

```{r}
download.file(url = "https://www.cbs.gov.il/he/mediarelease/doclib/2019/347/27_19_347t1.xls",
              destfile = "cbs_accidents.xls",
              mode = "wb")

download.file(url = "https://www.cbs.gov.il/he/publications/doclib/2019/1772/t01.xls",
              destfile = "cbs_km.xls",
              mode = "wb")

download.file(url = "https://www.cbs.gov.il/he/publications/doclib/2019/1762/t01.xls",
              destfile = "cbs_vehicles.xls",
              mode = "wb")
```

Load the data to R, with some wrangling to transform from human-readable to machine-readable:
```{r}
cbs_accidents <-read_excel(path = "cbs_accidents.xls",
                           col_names = c("year","fatalities","drop")
                           )%>%
  select(-drop) %>%
  mutate_all(as.numeric) %>%
  filter(!is.na(fatalities), !is.na(year))

cbs_km <- tibble(year = c(1975,1995,2000,2010:2018),
                 km_total = c(9232, 30633, 36482, 49870, 50576, 50107, 51207, 52400, 54820, 57220, 59602, 61196),
                 km_avg_percar = c(NA, 21.6, 20.5, 19.9, 19.1, 18.2, 18.1, 17.9, 17.9, 17.9, 17.7, 17.5)) 

cbs_vehicles <- read_excel(path = "cbs_vehicles.xls",
                           skip = 7) %>%
  rename(year = ...12, vehicles = total) %>%
  select(year, vehicles) %>%
  filter(!is.na(year)) %>%
  mutate(year2 = if_else(str_length(year) == 4, as.double(year), NA_real_),
         year2 = if_else(is.na(year2), lag(year2) -1, year2)) %>%
  select(-year, year = year2, vehicles) %>%
  filter(!is.na(vehicles), vehicles != 100, !is.na(year)) %>%
  mutate(vehicles = as.integer(vehicles))
```

Manually update fatalities with data for 2019 (https://www.cbs.gov.il/he/mediarelease/DocLib/2020/151/27_20_151b.pdf):
```{r}
cbs_accidents <- cbs_accidents %>%
  add_row(year = 2019, fatalities = 355)
```

Estimate KM traveled for 2019 (updated data not available as of today):
```{r}
km_fit <- lm(data = cbs_km %>% filter(year >= 2010),
             formula = km_total ~ year)

modelsummary::modelsummary(km_fit)

cbs_km <- cbs_km %>%
  add_row(year = 2019)

cbs_km$km_total_pred <- predict(km_fit, cbs_km)

cbs_km <- cbs_km %>% 
  mutate(km_total = if_else(year == 2019, round(km_total_pred), km_total ))
```

Manually update vehicles data for 2019 (https://www.cbs.gov.il/he/mediarelease/Pages/2020/כלי-רכב-מנועיים-בישראל-בשנת-2019.aspx):
```{r}
cbs_vehicles <- cbs_vehicles %>%
  add_row(year = 2019, vehicles = 3600600)
```

Load world bank data and PWT data on population in Israel:
```{r}
israel_pop1 <- WDI(country = "IL", indicator = c(pop = "SP.POP.TOTL"))

data("pwt9.1")

israel_pop2 <- pwt9.1 %>%
  filter(country == "Israel") %>%
  select(country, year, pop) %>%
  mutate(pop = pop * 10^6)

israel_pop <- full_join(israel_pop1, israel_pop2, by="year") %>%
  mutate(pop = if_else(!is.na(pop.x), pop.x, pop.y))
```

Merge to create rates then reshape to easily plot all variations:
```{r}
merged <- left_join(cbs_accidents, israel_pop, by="year") %>%
    left_join(cbs_km, by="year") %>%
  left_join(cbs_vehicles, by = "year") %>%
  mutate(rate = (fatalities/pop) * 10^5, 
         rate_1bkm = (fatalities/km_total) * 10^3,
         rate_100kcars = (fatalities/vehicles) * 10^5) %>%
  pivot_longer(cols = c("fatalities", "rate", "rate_1bkm", "rate_100kcars"), 
               names_to = "type", values_to = "value") %>%
  mutate(type = factor(type, 
                       levels = c("fatalities", "rate", "rate_1bkm", "rate_100kcars"),
                       labels =  c("הרוגים",
                                   "שיעור הרוגים ל-100 אלף איש",
                                   "הרוגים למיליארד קילומטר נסועה",
                                   "הרוגים ל-100 אלף כלי רכב"))
        ) %>%
  group_by(type) %>%
  mutate(max = max(value, na.rm=TRUE), min = min(value, na.rm=TRUE)) 
```


# Plotting

Plot absolute and rate of fatalities: 
```{r message=FALSE, warning=FALSE}

plot1 <- merged %>%
  filter(type %in% c("הרוגים",
                                   "שיעור הרוגים ל-100 אלף איש")) %>%
  ggplot(aes(x=year, y=value, color=type)) +
  geom_line(size = 1) +
  geom_hline(aes(yintercept = max), color="grey50", linetype="dashed") +
  geom_hline(aes(yintercept = min), color="grey50", linetype="dashed") +
  scale_y_continuous(breaks = pretty_breaks(n = 10)) +
  scale_x_continuous(breaks = pretty_breaks(n = 10)) +
  scale_color_manual(values = c("dodgerblue1","firebrick1"),
                     guide = FALSE) +
  facet_wrap(~type, scales = "free") +
  labs(x = "שנה",
       y = "",
       title = "תאונות דרכים בישראל: 2019-1949",
       subtitle = "קווים אופקיים מציגים את המקסימום והמינימום",
       caption = "מקור: נתוני הלשכה המרכזית לסטטיסטיקה")

plot1

ggsave(filename = "accidents_plot.png", width = 7)

```

Plot.ly for interactive chart:

```{r message=FALSE, warning=FALSE}
plotly<-ggplotly(plot1, tooltip = c("x","y"))
hide_legend(plotly)
```

Plotting per KM traveled:

```{r}
plot2 <- merged %>%
  filter(type == "הרוגים למיליארד קילומטר נסועה",
         !is.na(value)) %>%
  ggplot(aes(x=factor(year), y=value)) +
  geom_hline(aes(yintercept = max), color="grey50", linetype="dashed") +
  geom_hline(aes(yintercept = min), color="grey50", linetype="dashed") +
  geom_col(color = "dodgerblue1", fill="dodgerblue1", width = 0.8) +
  geom_label(aes(label = round(value, digits = 2)), color="dodgerblue1") +
  scale_y_continuous(breaks = pretty_breaks(n = 10), 
                     expand = expansion(mult = c(0, 0.1))) +
  labs(x = "שנה",
       y = "",
       title = "הרוגים למיליארד קילומטר נסועה בישראל: שנים נבחרות",
       subtitle = paste0("בשנת 2019 הנסועה נאמדה על בסיס הנסועה בין 2010 ל-2018", 
                        "\n",
                        "קווים אופקיים מציגים את המקסימום והמינימום"),
       caption = "מקור: נתוני הלשכה המרכזית לסטטיסטיקה")

plot2

ggsave(filename = "deaths_per1Bkm.png", width = 7)

```

Plotting per cars:

```{r}
plot3 <- merged %>%
  filter(type == "הרוגים ל-100 אלף כלי רכב",
         !is.na(value)) %>%
  ggplot(aes(x=year, y=value)) +
    geom_hline(aes(yintercept = max), color="grey50", linetype="dashed") +
  geom_hline(aes(yintercept = min), color="grey50", linetype="dashed") +
  geom_line(color = "dodgerblue1", size = 1) +
  scale_y_continuous(breaks = pretty_breaks(n = 10)) +
  scale_x_continuous(breaks = pretty_breaks(n = 10)) +
  labs(x = "שנה",
       y = "",
       title = "הרוגים ל-100 אלף כלי רכב בישראל: 2019-1965",
       subtitle = "קווים אופקיים מציגים את המקסימום והמינימום",
       caption = "מקור: נתוני הלשכה המרכזית לסטטיסטיקה")

plot3

ggsave(filename = "deaths_per1Kvehicles.png", width = 7)

```

Plotting per cars (focus on data from 1990):
```{r}
plot4 <- merged %>%
  filter(type == "הרוגים ל-100 אלף כלי רכב",
         !is.na(value),
         year >= 1990) %>%
  ggplot(aes(x=year, y=value)) +
  geom_line(color = "dodgerblue1", size = 1) +
  scale_y_continuous(breaks = pretty_breaks(n = 10)) +
  scale_x_continuous(breaks = pretty_breaks(n = 10)) +
  labs(x = "שנה",
       y = "",
       title = "הרוגים ל-100 אלף כלי רכב בישראל: 2019-1990",
       subtitle = "",
       caption = "מקור: נתוני הלשכה המרכזית לסטטיסטיקה")

plot4

ggsave(filename = "deaths_per1Kvehicles_from1990.png", width = 7)

```

